% ============================================================================
%  Federated and meta-learned blood-glucose forecasting across sites
%  PLOS Digital Health submission (PLOS template v3.8 Apr 2026)
%
%  STATUS: Full coherent draft. Abstract, Author summary, Introduction,
%  Materials and methods, Results (incl. finetuning), and Discussion (incl.
%  external-comparison subsection) written. Under reviewer-style revision.
%
%  NOTE: this uses the PLOS class options and the `cite` package (numeric
%  Vancouver citations). The bibliography style is `plos2015` (the genuine PLOS
%  Vancouver BibTeX style; the current template ships an identical file renamed
%  plos2025.bst). plos2015.bst is committed alongside this .tex so it compiles
%  as-is on Overleaf. The bib database is references.bib.
% ============================================================================
\documentclass[10pt,letterpaper]{article}
\usepackage[top=0.85in,left=2.75in,footskip=0.75in]{geometry}

\usepackage{amsmath,amssymb}
\usepackage{changepage}
\usepackage{textcomp,marvosym}
\usepackage{cite}
\usepackage{nameref,hyperref}
\usepackage[right]{lineno}
\usepackage[nopatch=eqnum]{microtype}
\DisableLigatures[f]{encoding = *, family = * }
\usepackage[table]{xcolor}
\usepackage{array}

\newcolumntype{+}{!{\vrule width 2pt}}
\newlength\savedwidth
\newcommand\thickcline[1]{%
  \noalign{\global\savedwidth\arrayrulewidth\global\arrayrulewidth 2pt}%
  \cline{#1}%
  \noalign{\vskip\arrayrulewidth}%
  \noalign{\global\arrayrulewidth\savedwidth}%
}
\newcommand\thickhline{\noalign{\global\savedwidth\arrayrulewidth\global\arrayrulewidth 2pt}%
\hline
\noalign{\global\arrayrulewidth\savedwidth}}

% Text layout
\raggedright
\setlength{\parindent}{0.5cm}
\textwidth 5.25in
\textheight 8.75in

\usepackage[aboveskip=1pt,labelfont=bf,labelsep=period,justification=raggedright,singlelinecheck=off]{caption}
\renewcommand{\figurename}{Fig}

\bibliographystyle{plos2015}

\makeatletter
\renewcommand{\@biblabel}[1]{\quad#1.}
\makeatother

% Header and Footer with logo
\usepackage{lastpage,fancyhdr,graphicx}
\usepackage{epstopdf}
\pagestyle{fancy}
\fancyhf{}
\rfoot{\thepage/\pageref{LastPage}}
\renewcommand{\headrulewidth}{0pt}
\renewcommand{\footrule}{\hrule height 2pt \vspace{2mm}}
\fancyheadoffset[L]{2.25in}
\fancyfootoffset[L]{2.25in}
\lfoot{\today}

\begin{document}
\vspace*{0.2in}

% Title must be 250 characters or less. Sentence case.
\begin{flushleft}
{\Large
\textbf{Federated and meta-learned blood-glucose forecasting across sites: a domain-generalisation benchmark and a cross-site deployment prototype}
}
\newline
\\
Name1 Surname\textsuperscript{1*}
\\
\bigskip
\textbf{1} Affiliation Dept/Program/Center, Institution Name, City, State, Country
\\
\bigskip

* correspondingauthor@institute.edu

\end{flushleft}

% Please keep the abstract below 300 words
\section*{Abstract}
Algorithms that predict blood glucose 30 minutes ahead are increasingly used in
type~1 diabetes (T1D) to warn of excursions and support insulin decisions, but a
model trained on one cohort often transfers poorly to new patients, devices, and
sites --- and the diverse, multi-site data that would make it robust are kept
siloed by privacy regulation. Federated learning (FL) lets institutions train a
shared model without moving raw records. Yet for glucose forecasting the leading
federated strategies have rarely been compared head-to-head, evaluations seldom
separate accuracy on the training cohorts (held-in) from accuracy on entirely
unseen cohorts (out-of-distribution, OOD), and almost all such work is simulated
rather than run across real machines. We address these gaps. Using a common
encoder--decoder transformer and four real T1D cohorts as federated clients ---
with a fifth, never-seen cohort (ReplaceBG) as an OOD test set --- we benchmark
single-cohort local training against federated averaging (FedAvg), FedProx,
meta-learning for domain generalisation (MLDG), and two personalised-FL methods,
under a strict by-patient split and a five-seed protocol, reporting held-in and
OOD error separately at the 30-minute horizon. Federation matched local training
in-domain (held-in average $\approx$20 mg/dL RMSE) and transferred more reliably
out of it: the global models reached $\approx$21.3--21.6 mg/dL on the unseen
cohort, whereas single-cohort models were more variable --- one (trained on the
smallest cohort) transferred markedly worse. The gains over local training are
small relative to CGM sensor error, so the practical value is reduced
deployment risk rather than a large accuracy jump; the federated strategies were
statistically indistinguishable from one another. Finetuning the federated model
on a new site did slightly better than training there from scratch. To show the
approach is not merely a simulation, we built a lightweight FL system --- a
stateless server plus per-cohort clients communicating over HTTP with only the
Python standard library --- and ran it end-to-end across two physically separate
machines over the public internet. Federation thus offers T1D forecasting a way
to reduce the risk of catastrophic transfer failure while keeping raw data
in-house, using tooling light enough for a small clinical consortium.

% Author Summary: 150-200 words, first person.
\section*{Author summary}
People with type~1 diabetes rely increasingly on algorithms that predict where
their glucose will be in the next half hour. These predictors work well for the
patients they were trained on but can fail when applied to a different clinic,
device, or population --- and the varied data that would make them robust are
scattered across institutions that, for privacy reasons, cannot pool raw
records. Federated learning lets those institutions train a shared model while
each keeps its own data in-house. We asked which federated method to use, and
whether the approach works outside a simulation. Treating four separate
diabetes studies as collaborating sites, we compared the main federated
strategies and tested each on a fifth study that contributed no training data.
We found that federation was about as accurate as training on a single cohort
for patients from the participating sites, and more reliable on the unseen
cohort: a model trained on one cohort alone sometimes failed badly on new
patients, whereas the federated models did not. The accuracy differences are
small, so the main benefit is lower risk of a bad surprise on a new population
rather than a large accuracy gain. We also built a minimal system and ran the
training for real across two separate computers over the internet. Our results
give a small group of clinics a practical, privacy-preserving starting point for
building glucose predictors that are less likely to fail on patients they have
never seen --- though clinical metrics beyond error size, and tests on more
diverse populations, are still needed before deployment.

\clearpage
\newgeometry{top=0.85in,left=1in,right=1in,footskip=0.75in}
\linenumbers

% ===========================================================================
\section*{Introduction}

Type~1 diabetes (T1D) affects an estimated 8.4~million people worldwide and its
incidence is rising in many regions~\cite{idf2021,gregory2022t1dindex}. People living with T1D must
continually balance insulin, food, and activity to keep blood glucose within a
narrow target range; sustained excursions above or below that range drive both
the acute dangers of hypoglycaemia and the long-term microvascular and
cardiovascular complications of hyperglycaemia. Continuous glucose monitoring
(CGM), which reports interstitial glucose roughly every five minutes, is now
central to diabetes care; it underpins clinical outcomes such as time in range,
for which international consensus targets have been defined~\cite{battelino2019tir}. The same dense data stream makes
short-horizon glucose \emph{forecasting} clinically valuable: a reliable
prediction of where glucose will be in 30--60 minutes can trigger preemptive
alerts, inform bolus and snack decisions, and close the loop in automated
insulin-delivery systems~\cite{vettoretti2020advanced,bekiari2018ap}.

Short-horizon CGM forecasting has progressed from autoregressive and classical
machine-learning models to deep sequence models. Recurrent architectures with
predictive variance estimation established strong baselines on benchmark
cohorts~\cite{martinsson2020}, and standardised datasets and challenges shaped
much of the early evaluation~\cite{marling2020ohio}. The
transformer~\cite{vaswani2017attention} and its long-horizon time-series
variants~\cite{zhou2021informer,wu2021autoformer} have since been adapted to
glucose, with personalised, uncertainty-aware transformer forecasters reporting
state-of-the-art accuracy~\cite{sergazinov2023gluformer,sergazinov2024glucobench}.
These models are accurate when trained and evaluated on the same cohort, but how
well they transfer across cohorts is an open question~\cite{ghimire2024generalize}:
glucose dynamics, demographics, and devices differ from site to site, and a model
can degrade on populations it did not see in training~\cite{farahmand2025glumind}.
A model that has only seen one cohort therefore has little reason to generalise to
the next; robust generalisation instead calls for training over large, diverse
populations spanning many sites.

Large, diverse training data is exactly what is hardest to assemble. CGM traces
are sensitive personal health records, fragmented across hospitals, clinical
trials, device manufacturers, and national registries, and governed by
data-protection regimes such as the GDPR and HIPAA that restrict moving raw
records between institutions, let alone across borders. The data that improve
generalisation and the data that exist are thus mismatched: large, heterogeneous
corpora are what is needed, but what is available is many small, governed silos.
Federated learning (FL) addresses this tension~\cite{mcmahan2017fedavg}: sites
train on their own data and exchange only model updates, so a shared model is
learned without any raw record leaving its institution. On this basis FL has been
widely adopted for collaborative medical machine
learning~\cite{rieke2020future,sheller2020federated,kaissis2020secure}. Its
central difficulty is statistical heterogeneity: plain federated averaging
degrades when client distributions diverge, motivating proximal
regularisation~\cite{li2020fedprox} and control-variate variance
reduction~\cite{karimireddy2020scaffold}. Most of this work, however, is evaluated
in simulation, where ``clients'' are partitions of one dataset on a single
machine.

Even with federation, real CGM cohorts differ in device generation, demographics,
treatment protocols, and glycaemic regime, and these differences widen across
countries. A federated model must therefore cope with substantial domain shift: it
should generalise to patients, and even whole cohorts, it never trained on (domain
generalisation), and/or specialise to each site (personalisation). Domain
generalisation seeks models that transfer to unseen
distributions~\cite{zhou2022dgsurvey}, and meta-learning for domain generalisation
(MLDG) simulates train/test domain shift within each update so the model learns to
generalise rather than memorise~\cite{li2018mldg}. Bringing these ideas into
federation is an active area surveyed by Li et al.~\cite{li2025fedDGsurvey},
spanning several families. Meta-learning approaches adapt the model-agnostic
meta-learning objective to federation, for communication-efficient
convergence~\cite{chen2018fedmeta} or for models that personalise to a new client
in a few gradient steps~\cite{fallah2020perfedavg}. Personalisation approaches keep
a per-client model: APFL learns an adaptive convex mixture of the global and a
local model~\cite{deng2020apfl}, while Ditto regularises each client's personal
model toward the global one with a tunable proximal strength~\cite{li2021ditto}.
Other approaches instead align client representations, via model-contrastive
objectives~\cite{li2021moon}, federated contrastive learning for multi-centre
medical imaging~\cite{fedcl2023}, or adversarial domain
alignment~\cite{flda2023,fedadv2022}. Our benchmark covers the first two
families---meta-learning (MLDG) and personalisation (APFL, Ditto)---alongside
standard federated averaging (FedAvg, FedProx) and single-cohort local training,
and leaves representation-alignment methods to future work.

\textbf{Yet these methods are rarely compared head-to-head for glucose
forecasting.} They are seldom benchmarked against one another on a common task, and
evaluations rarely separate the two regimes that matter clinically---accuracy on
patients from the training cohorts (held-in) versus accuracy on entirely unseen
cohorts (out of distribution, OOD)---even though a method can behave quite
differently in the two. Closing this gap---comparing federated strategies on one
glucose task under a shared protocol, with held-in and OOD performance reported
separately---is the main focus of this paper.

A second, more practical observation is that almost all federated glucose work is
simulated rather than deployed, with ``clients'' as partitions of one dataset on a
single machine. Running a federation for real additionally requires networked,
physically separate sites to exchange updates and to start from a common model;
production FL frameworks can do this but carry substantial dependencies and
assumptions that can be ill-suited to a small clinical consortium that simply wants
a handful of sites, one cohort each, to train together. We therefore also ask a
feasibility question: whether a lightweight cross-site federation for CGM
forecasting can be built and run across real machines with minimal tooling.

This paper makes one primary contribution and one secondary one. \textbf{Our
primary contribution is a controlled domain-generalisation benchmark of federated
strategies for transformer-based glucose forecasting.} Using a common
encoder--decoder transformer and four real adult T1D cohorts as federated
clients, with a held-out cohort as an OOD test site, we compare single-cohort
local training, FedAvg, FedProx, MLDG, and two personalised-FL methods (APFL and
Ditto) under a multi-seed protocol, reporting held-in and OOD accuracy separately
at the 30-minute horizon --- a horizon commonly used for preemptive glucose
alerts. All cohorts are adult T1D populations, so our findings should not be
assumed to extend to paediatric or type-2 diabetes.

\textbf{As a secondary contribution, we show that such a federation can be run for
real, not only simulated.} We build, from scratch, a lightweight cross-site FL
system for CGM forecasting: a stateless aggregation server that holds the global
model but no data, and clients that each wrap one institution's cohort,
communicating over HTTP using only the Python standard library. The server
broadcasts the initial weights so every client starts from an identical model, and
the same protocol carries FedAvg, FedProx, and MLDG training. We validate it
end-to-end across two physically separate machines connected over the public
internet as a feasibility prototype, and describe how the design could extend to
a geographically distributed, multi-country deployment. We do not claim a
production-grade or multi-country deployment here: establishing that would
require secure aggregation, encryption, and more than two independent sites,
which we leave to future work.

Taken together, the benchmark shows how to evaluate federated glucose
forecasting honestly and what such an evaluation reveals, while the prototype
shows that a federation of this kind can be run across physically separate
machines, not only simulated.

% ===========================================================================
\section*{Materials and methods}

\subsection*{Datasets and federation setup}

% REVIEW (citation check 2026-06-19): Table~\ref{tab:datasets} patient counts
% disagree with the source dataset papers -- HUPA-UCM is 22 here vs 25 in
% Hidalgo et al. 2024; T1D-UOM is 14 here vs 17 in Alsuhaymi et al. 2025 (21
% enrolled, 4 withdrew). These are presumably our usable post-filtering subsets;
% either reconcile the numbers or state the filtering criterion explicitly.
We use five publicly available type-1 diabetes (T1D) continuous glucose
monitoring (CGM) datasets. Four serve as the federated clients --- HUPA-UCM
\cite{hidalgo2024hupaucm}, ABC4D, ARISES, and T1D-UOM
\cite{alsuhaymi2025t1duom} --- with \emph{one dataset per client}. Each dataset
originates from a distinct study, population, and device, so a per-dataset
partition makes every client a genuinely different domain rather than an
arbitrary shard of a pooled corpus; this is what lets the federation function as
a test of cross-domain generalisation. The fifth dataset, ReplaceBG
\cite{aleppo2017replacebg}, is never seen during training and is used only as a
held-out, out-of-distribution (OOD) test client to measure how well a federated model transfers to a cohort that
contributed no data to training. Within each client, we partition patients
into disjoint train, validation, and test groups, splitting strictly by
patient ID so that no patient contributes data to more than one group. Every
window in the validation and test sets therefore comes from a patient never
seen during training, and reported performance measures generalisation to
unseen patients rather than to unseen windows of patients already observed.

Two of the four client cohorts, HUPA-UCM and T1D-UOM, together with the held-out
ReplaceBG cohort, are obtained through MetaboNet
\cite{wolff2026metabonet}, a recently released resource that consolidates many
public T1D datasets into a single harmonised schema with a common 5-minute grid
and unified CGM, insulin, and carbohydrate fields. Drawing these cohorts from
MetaboNet ensures their variables are already aligned to a shared format; the
remaining two cohorts, ABC4D and ARISES, are converted into the same schema
from their original releases. We do not adopt MetaboNet's own train/test split,
which allows the same patient to appear in both partitions; instead we pool the
data and re-split strictly by patient.

Table~\ref{tab:datasets} summarises each cohort. Mean and standard deviation of
CGM are reported in mg/dL, and time-in-range (TIR) is the fraction of CGM
readings in the clinical 70--180 mg/dL band.

\begin{table}[!ht]
\centering
\captionsetup{justification=centering,singlelinecheck=on}
\caption{\bf Federated client cohorts.}
\label{tab:datasets}
\begin{tabular}{lrrrrr}
\thickhline
dataset & patients & total windows & mean CGM & sd CGM & TIR (\%) \\
        &          &               & (mg/dL)  & (mg/dL) & 70--180  \\
\hline
HUPA-UCM & 22 & 291,728 & 140.9 & 56.6 & 72.2 \\
ABC4D    & 25 & 562,043 & 155.1 & 63.4 & 65.5 \\
ARISES   & 12 & 126,248 & 160.5 & 63.7 & 63.7 \\
T1D-UOM  & 14 & 300,796 & 151.6 & 59.9 & 70.5 \\
\thickhline
\end{tabular}
\begin{center}
{\small ReplaceBG (OOD test client) is excluded from training and reported
separately. ARISES is both the highest-glucose and the smallest cohort.}
\end{center}
\end{table}

\subsection*{Preprocessing and windowing}

Every cohort is resampled to a uniform 5-minute grid. We form supervised
examples with a sliding window of 72 samples (6\,h) of CGM context to forecast
the next 12 samples (60\,min), advancing the window one sample at a time
(stride~1). Windows that would bridge a sensor gap longer than 6 samples
(30\,min) are discarded, so every example is contiguous. CGM is normalised with
a single \emph{global} mean and standard deviation ($154.04 \pm 61.00$ mg/dL)
computed from the four training cohorts only --- the held-out OOD cohort does not
contribute to it, so no OOD information enters preprocessing --- and shared
across all clients rather than using per-client statistics. A common
normalisation keeps the global model's input scale identical at every site,
which is what allows weights aggregated across clients to be applied directly to
the OOD cohort; it is shipped with the model as a fixed constant. Alongside CGM,
each example carries a time-of-day mark encoded
as $(\sin 2\pi\,\mathrm{frac},\ \cos 2\pi\,\mathrm{frac})$, where $\mathrm{frac}$
is the fraction of the day, to expose circadian structure.

\subsection*{Forecasting model}

The forecaster is a sequence-to-sequence (encoder--decoder) Transformer that maps
the 72-step CGM context to the 12-step horizon. It uses
$d_{\text{model}}=128$, 8 attention heads, 6 encoder layers, 2 decoder layers,
feed-forward width $d_{\text{ff}}=2048$, ReLU activations, and dropout~$0.05$
($\approx 4.9$\,M parameters). Decoding is single-shot rather than
autoregressive: the decoder is fed a learned start token while the encoder
carries the full history, and the 12-step horizon is produced in one forward
pass. The decoder predicts a single CGM channel ($c_{\text{out}}=1$) and the
model is trained with mean-squared-error (MSE) loss. The same architecture is
used for every client and every federated strategy, so all comparisons differ
only in how the model is trained, not in capacity. The model emits a point
forecast only; it produces no predictive-uncertainty estimate (see Limitations).

\paragraph{Bolus input variant}
As a secondary analysis, we test whether adding an exogenous input improves
forecasting. We train a two-channel variant ($\texttt{enc\_in}=2$) in which a
second encoder channel carries $\log(1+\mathrm{bolus}_{\mathrm{IU}})$, the
log-transformed bolus insulin dose; bolus events are sparse (roughly 1\% of
timesteps are non-zero). The decoder remains CGM-only. Because bolus fields are
harmonised only across a subset of the cohorts and the effect proves small
relative to seed-to-seed variation, we treat this as a preliminary sensitivity
check and report it as a delta against the CGM-only model rather than as a main
result.

\subsection*{Federated learning system}

Training is coordinated by a from-scratch, networked federated learning system
in which one institution runs a server and each cohort runs a client. The server
holds the global model weights but never sees any data: at round one it mints a
single seeded set of initial weights and broadcasts them, so every client begins
from byte-identical parameters regardless of hardware. Each round, clients fetch
the current global weights, train locally on their own cohort, and return updated
weights; the server aggregates them by FedAvg and advances to the next round.
Clients communicate with the server over plain HTTP through a small task
state-machine (train\,$\rightarrow$\,validate\,$\rightarrow$\,test), so the same
protocol carries FedAvg, FedProx, and MLDG without modification --- the choice of
strategy is entirely client-side, and the server always performs the same FedAvg
aggregation. We validate the system end-to-end across two physically separate
machines communicating over the public internet as a feasibility prototype, and
describe how the design could generalise to a geographically distributed,
multi-country deployment. The prototype uses plain HTTP without encryption,
secure aggregation, or differential privacy; it demonstrates the training
workflow, not a production-grade secure channel (see Limitations).

\subsection*{Federated strategies and baselines}

\textbf{Local (baseline).} Our primary comparator is local training: a separate
model trained on a single cohort's data only, with no federation. The central
question of the benchmark is whether federation improves on this baseline ---
both when evaluated on the same cohorts that participated (held-in) and on the
unseen OOD cohort. For OOD we report the mean over the four single-cohort models.

\textbf{FedAvg / FedProx.} Standard federated averaging, and FedProx, which adds
a proximal term $(\mu/2)\lVert w - w_{\text{global}}\rVert^2$ to each client's
objective to limit client drift. We run FedProx at a single proximal strength
($\mu=0.05$) and did not sweep it, so its ranking relative to FedAvg should be
read as indicative rather than as a tuned comparison.

\textbf{MLDG.} Meta-learning for domain generalisation, adapted to the federated
setting: at each client step the client's patient groups are split into
meta-train and meta-test partitions, an inner adaptation step is taken on
meta-train, and the model is updated by the gradient of the meta-test loss
evaluated after that step. Because the split is over patients \emph{within} a
client, MLDG here simulates cross-\emph{patient} shift inside a cohort, not the
cross-\emph{cohort} shift measured by the OOD test. MLDG and FedProx alter only
client-side training; the server still FedAvg-aggregates. The precise inner/outer
update (including the inner learning rate and the treatment of the optimiser
state) is defined in the released code.

\textbf{Personalised FL (PFL).} APFL learns a per-client mixture
$\alpha v_i + (1-\alpha)w$ of a personal model $v_i$ and the global model $w$
with an adaptively updated $\alpha$ (initial $0.25$); we also include a decoupled
APFL variant. Ditto trains a personal model pulled toward the global one by a
proximal term ($\mu\in\{0.01,0.1,1.0\}$). These methods produce one personalised
model per client and \emph{no} single global model, so --- like the local
baseline's per-cohort models --- they are evaluated held-in only and cannot be
applied to the OOD cohort. We flag that the APFL mixing weight may not have moved
far from its initial value in our runs; its numbers should therefore be read as a
lower bound on what a fully tuned APFL could achieve.

\subsection*{Evaluation protocol and metrics}

We evaluate every strategy in two regimes. \emph{Held-in} performance is measured
on each participating cohort's own held-out patients, isolating predictive
accuracy at the contributing sites. \emph{OOD} performance is measured on
ReplaceBG, which contributed no training data, isolating cross-domain transfer;
only strategies that yield a single deployable global model (Local-mean, FedAvg,
FedProx, MLDG) receive an OOD score. We report root-mean-squared error (RMSE) in
mg/dL at the 30-minute horizon; the model forecasts to 60 minutes, but we defer
the longer-horizon analysis to future work. All runs use early stopping on
average validation MSE (patience~5 over a maximum of 25 communication rounds),
and every configuration is repeated over 5 random seeds (42, 43, 44, 46, 47; seed
45 was not run). We report the mean and standard deviation across seeds. A few
cells are computed over 4 rather than 5 seeds because a run is missing (the OOD
runs for FedAvg and FedProx at one seed; one T1D-UOM seed), so the OOD comparison
between MLDG and FedAvg/FedProx is not over an identical seed set. We emphasise
that these standard deviations quantify \emph{optimisation stochasticity} for one
fixed patient split; because we use a single by-patient split, they understate
the patient-sampling uncertainty that would matter at deployment, and should not
be read as precision on new patients. We report a single aggregate error metric
(RMSE); clinically oriented metrics (an error-grid analysis, MARD, and
hypo-/hyperglycaemia event detection) and predictive uncertainty are not included
here and are discussed as limitations.

\subsection*{Implementation and training details}

All models are trained with the Adam optimiser (learning rate~$10^{-4}$, weight
decay~$10^{-3}$, batch size~256, one local epoch per communication round). The
federated system is implemented in Python using only the standard library for
networking.

\subsection*{Data and ethics}

All five cohorts are publicly available, de-identified secondary datasets, used
here under their respective terms of access; no new human-subjects data were
collected for this study. HUPA-UCM \cite{hidalgo2024hupaucm}, T1D-UOM
\cite{alsuhaymi2025t1duom}, and ReplaceBG \cite{aleppo2017replacebg} are obtained
via MetaboNet \cite{wolff2026metabonet}; ABC4D and ARISES are used from their
original releases. Code for the federated system and the analysis is available
from the authors.
% NOTE (author): add a public repository/DOI link and confirm each dataset's
% access route and licence before submission; add an FL-architecture figure.

\section*{Results}

Table~\ref{tab:main} reports held-in RMSE on each client's unseen test patients
and OOD RMSE on ReplaceBG, at the 30-minute horizon, for the CGM-only model.

\begin{table}[!ht]
\centering
\captionsetup{justification=centering,singlelinecheck=on}
\caption{\bf CGM-only RMSE (mg/dL) at the 30-minute horizon.}
\label{tab:main}
\begin{tabular}{lrrrrrr}
\thickhline
 & \multicolumn{5}{c}{held-in (per client)} & OOD \\
\cline{2-6}
strategy & HUPA-UCM & ABC4D & ARISES & T1D-UOM & avg & ReplaceBG \\
\hline
Local (per cohort) & 18.81 & 19.95 & 22.04 & 19.80 & 20.15 & 23.54 \\
\hline
FedAvg  & 18.64 & 19.80 & 22.38 & 19.53 & 20.09 & 21.43 \\
FedProx & 18.89 & 19.94 & 22.60 & 19.67 & 20.28 & 21.57 \\
MLDG    & \textbf{18.46} & \textbf{19.68} & 22.30 & 19.53 & \textbf{19.99} & \textbf{21.34} \\
\hline
APFL          & 19.11 & 19.88 & 22.53 & 19.61 & 20.28 & --- \\
APFL-decoupled & 19.37 & 19.90 & 22.47 & 19.73 & 20.37 & --- \\
Ditto ($\mu{=}0.01$) & 18.90 & 19.90 & \textbf{21.98} & 19.64 & 20.11 & --- \\
Ditto ($\mu{=}0.1$)  & 18.90 & 19.93 & 22.28 & 19.66 & 20.19 & --- \\
Ditto ($\mu{=}1.0$)  & 18.74 & 19.77 & 22.35 & \textbf{19.50} & 20.09 & --- \\
\thickhline
\end{tabular}
\begin{center}
{\small Mean over 5 random seeds; a few cells use 4 seeds where a run is missing
(notably the FedAvg and FedProx OOD columns), so the OOD column does not compare
an identical seed set across rows. Standard deviations (omitted here for space)
are across seeds and are small ($\approx 0.1$--$0.2$ mg/dL held-in), except the
Local OOD entry, which is noisy across seeds (sd $\approx 1.65$ mg/dL). \emph{avg}
is the mean of the four held-in client means. Local held-in is each cohort's own
single-cohort model; Local OOD (23.54) is the mean of the four single-cohort
models on ReplaceBG. Personalised strategies (APFL, Ditto) yield no single
global model and so receive no OOD score. Bold marks the numerically lowest value
per column; differences among the federated strategies are within seed variation
(see text).}
\end{center}
\end{table}

\paragraph{Federation matches local accuracy held-in and transfers more reliably
out-of-distribution.} On the contributing cohorts, the global FedAvg and MLDG
models are on par with cohort-specific local training (held-in average $20.09$
and $19.99$ vs.\ $20.15$ mg/dL): pooling four heterogeneous cohorts through
federation does not cost in-domain accuracy despite each client seeing only the
shared global weights. The difference emerges out-of-distribution, and it is
about reliability more than average accuracy. On the never-seen ReplaceBG cohort
the federated models scored $21.34$--$21.57$ mg/dL against a mean single-cohort
model of $23.54$ mg/dL, but that mean is dominated by one failure: three of the
four single-cohort models transferred to within roughly half a mg/dL of the
federated models, while the ARISES-only model collapsed on ReplaceBG
($28.51 \pm 6.85$ mg/dL) and drove the local baseline's high seed-to-seed
variance (sd $\approx 1.65$ mg/dL). The honest reading is therefore not that
federation is $2$ mg/dL more accurate --- a gap that is in any case small
relative to CGM sensor error ($\approx 8$--$15$ mg/dL) --- but that a
single-cohort model can occasionally transfer catastrophically and one cannot
know in advance which cohort's model will fail, whereas none of the federated
models did so. Federation's value here is reduced tail risk, not a large mean
gain.

\paragraph{The federated strategies are statistically indistinguishable from one
another.} Among the strategies that produce a deployable global model, MLDG has
the numerically lowest held-in average and OOD error and is numerically best on
two of the four clients, but the margins ($\approx 0.1$ mg/dL held-in and OOD)
are smaller than the across-seed standard deviations, we ran no significance
test, and the OOD comparison mixes 5-seed and 4-seed means (Table~\ref{tab:main}
caption). We therefore do not claim MLDG is the best method; the defensible
statement is that all federated strategies transfer comparably and all avoid the
single-cohort failure mode. FedProx was numerically lowest here but was run at a
single untuned $\mu$, so this is not a tuned ranking. The personalised methods do
not change the held-in picture --- their averages cluster around FedAvg's
($20.1$--$20.4$) --- and by construction cannot be deployed on the OOD cohort.

\paragraph{A federated model is a competitive starting point for a new site.} We
next asked whether the federated model helps a \emph{new} site that does
contribute data, by finetuning the global model on ReplaceBG's own training
patients and comparing against a model trained on ReplaceBG from scratch.
Finetuning any of the three global models reached a slightly lower RMSE on
ReplaceBG's held-out patients (FedProx$\rightarrow$finetune $20.98 \pm 0.06$,
FedAvg$\rightarrow$finetune $21.04 \pm 0.04$, MLDG$\rightarrow$finetune
$21.04 \pm 0.07$ mg/dL) than training on ReplaceBG alone
($21.38 \pm 0.09$ mg/dL). The differences are small ($\approx 0.3$--$0.4$ mg/dL)
and we report no significance test, so we read this as suggestive: the federated
model is at least a competitive initialisation, and training on the new site's
own data did not beat it. We note two caveats that bound this result: ReplaceBG
is a large cohort ($166$ training patients), so this does not test the
data-poor small-clinic regime the deployment argument targets, and the
from-scratch comparator used the same training budget, which we did not
separately tune.

\section*{Discussion}

\paragraph{Principal findings.} Across four real adult T1D cohorts federated as
distinct domains, with a fifth never-seen cohort as an out-of-distribution (OOD)
test, federation matched single-cohort local training in-domain and transferred
more reliably out of it. Held-in, the global FedAvg and MLDG models were on par
with cohort-specific local training (average RMSE $20.09$ and $19.99$ vs.\
$20.15$ mg/dL at the 30-minute horizon); OOD, the federated models
($\approx 21.3$--$21.6$ mg/dL) avoided the occasional catastrophic transfer
failure of single-cohort training, though most single-cohort models transferred
comparably and the mean gain is small relative to sensor error. The federated
strategies were statistically indistinguishable from one another. Beyond the
benchmark, we showed that a federation of this kind runs across physically
separate machines: the same protocol trains FedAvg, FedProx, and MLDG over the
public internet, with no raw data leaving any site.

\paragraph{Federation buys more reliable transfer, not only an accuracy gain.}
The clearest message of the benchmark is not that federation is more accurate
in-domain --- the held-in gap over local training is small --- but that it
lowers the risk of a catastrophic transfer failure that single-cohort training
carries. A model trained on one cohort can transfer badly: the ARISES-only
model, accurate on its own patients ($22.04$ mg/dL), collapsed on ReplaceBG
($28.51 \pm 6.85$ mg/dL) and drove most of the local baseline's high OOD variance
(sd $\approx 1.65$ mg/dL), even though the other three single-cohort models
transferred within about half a mg/dL of the federated models. The difficulty is
that one cannot know in advance which site's model will be the one that fails. On
the single held-out cohort we tested, every federated model avoided this failure.
We are careful not to over-read one OOD cohort and 4--5 seeds as a guarantee ---
a larger panel of held-out cohorts is needed to establish how general this is
(see Limitations) --- but for a clinical setting in which distribution shift is
inevitable, reducing the chance of a silent, cohort-specific collapse is a
practical argument for federation independent of any in-domain accuracy
difference.

\paragraph{Generalisation versus personalisation, and how to have both.} Our
results expose a genuine tension. The personalised strategies (Ditto, APFL) and
dedicated single-cohort models were the \emph{only} approaches that beat the
federated models on the hardest and smallest in-domain cohort: on ARISES, Ditto
($\mu{=}0.01$, $21.98$ mg/dL) and the dedicated ARISES model ($22.04$ mg/dL) edged
MLDG ($22.30$). Yet these same methods produce no single global model and so
cannot serve the OOD cohort at all --- exactly the regime in which the global
models are most valuable. The two objectives pull in opposite directions:
specialising to a site helps that site but forfeits transfer; generalising across
sites transfers but leaves per-site accuracy on the table. The way to obtain both
is to treat the generalised global model as a foundation and adapt it locally,
rather than choosing one regime outright.

\paragraph{Finetuning a federated model for a new site.} The finetuning
experiment points, tentatively, toward a ``foundation-then-adapt'' recipe.
Finetuning the global federated model on ReplaceBG's own patients reached
$20.98$--$21.04$ mg/dL, slightly better than either the zero-shot federated model
($21.34$ mg/dL) or a model trained on ReplaceBG from scratch ($21.38$ mg/dL). The
differences are small ($\approx 0.3$--$0.4$ mg/dL) and untested for significance,
so we draw only qualitative conclusions. First, training on the site's own data
in isolation did not beat the finetuned federated model, suggesting the shared
model contributes usefully. Second, the three global strategies converge once
finetuned (FedProx, numerically lowest zero-shot here, is also numerically best
after finetuning), consistent with the choice of federated strategy mattering
most for zero-shot transfer and mattering less once local data are available. A
plausible practical protocol is therefore two-stage --- federate to obtain a
shared model, then finetune per site --- but establishing that a finetuned model
reliably beats local-only training, especially for a data-poor new site, requires
a data-efficiency study we have not yet run (this experiment used ReplaceBG's
large 166-patient training set).

\paragraph{On meta-learning.} MLDG was numerically best on average, but the
margin over plain FedAvg is within seed noise, so we do not claim a meta-learning
advantage. It is also worth noting a mechanistic caveat: our MLDG splits patients
into meta-train and meta-test \emph{within} each client, so it simulates
cross-\emph{patient} shift inside a cohort, not the cross-\emph{cohort} shift the
OOD test measures. A federated-DG formulation that meta-splits \emph{across}
clients would target the evaluated shift more directly and is a natural next
step. The dominant effect in our results is federation itself; the choice of
federated strategy appears second-order, and on the present evidence no single
strategy is clearly preferable.

\paragraph{Implications for a clinical consortium.} These findings give a small
group of institutions a reason to consider federating. A participating centre
loses little in-domain, and gains a model less likely to fail on a population it
did not train on. Because no raw record leaves any institution, federation
removes the raw-data-pooling barrier that data-protection regimes such as GDPR
and HIPAA impose --- though we stress that keeping raw data in-house is a
\emph{necessary but not sufficient} condition for compliance: a real deployment
would still need encryption in transit, secure aggregation or differential
privacy against inversion of the exchanged updates, access control, and an
appropriate governance and impact assessment, none of which our prototype
provides. Subject to those safeguards, the tooling itself is light enough to run
without dedicated federated-learning infrastructure. The benefit should also
compound as more cohorts join and the shared model becomes more representative.

\paragraph{Strengths.} Three features distinguish this study from prior federated
glucose work. First, it compares local training, FedAvg, FedProx, MLDG, and two
personalised-FL families head-to-head on one task under a shared protocol, rather
than reporting a single method in isolation. Second, it separates held-in from
OOD performance, a distinction that prior evaluations rarely make explicit but
that proves informative: the strategies are nearly indistinguishable in-domain,
and OOD the meaningful split is between single-cohort training (which can fail)
and federation (which did not, here), rather than between federated strategies.
Third, the cohorts are partitioned strictly by patient and normalised with a
single shared statistic, and the federation is run across real, physically
separate machines rather than simulated as shards of one dataset on one host.

\paragraph{Positioning against prior results on these cohorts.} Our per-cohort
accuracy is competitive with previously published results on the same datasets,
which is notable because we evaluate under a strictly harder protocol. The
distinction is the train/test split. Almost all prior work splits each
patient's series chronologically --- the first portion for training, the last
portion for testing (a \emph{by-time} split) --- so every test window comes from
a patient the model has already seen; the model need only extrapolate forward in
time for a familiar individual. We instead split \emph{by patient}: test patients
are disjoint from training patients, so every reported number measures
generalisation to people the model has never seen. By-patient evaluation is the
harder task and the one that matches deployment, where a predictor meets new
patients, but it typically yields higher error than a by-time split on the same
cohort. Table~\ref{tab:prior} collects the closest comparable external results.
On HUPA-UCM and T1D-UOM, the strongest real-data RMSE@30 baselines are from
GlucoFM-Bench \cite{lu2026glucofmbench} (a benchmark of time-series foundation
models, univariate CGM, by-time split): HUPA-UCM 16.71 and T1D-UOM 20.81 mg/dL.
On ARISES and ABC4D they are from the personalised evidential meta-learning
forecaster of Zhu et al.\ \cite{zhu2023fcnn} (CGM plus carbohydrate and insulin,
by-time split): 20.23 and 20.25 mg/dL. Our by-patient figures for a global model
(MLDG, numerically the best of our federated strategies; HUPA-UCM 18.46, ABC4D
19.68, ARISES 22.30, T1D-UOM 19.53 mg/dL) sit in the same 17--22 mg/dL range
despite the harder split and, on
HUPA-UCM/T1D-UOM, despite using CGM alone against models that also consume
insulin and carbohydrate inputs. We therefore do not claim to beat these numbers
--- a by-patient result is not expected to --- but to match them under an
evaluation that better reflects deployment. To our knowledge no prior paper
reports all four of these cohorts, separates held-in from out-of-distribution
performance, or evaluates across-patient generalisation as we do; the one prior
work that also splits by patient (AEGIS \cite{pansheriya2026aegis}, on T1D-UOM
via leave-one-patient-out) forecasts only to 15 minutes and so is not comparable
at our 30-minute horizon.

\begin{table}[!ht]
\centering
\captionsetup{justification=centering,singlelinecheck=on}
\caption{\bf Published RMSE@30 on our cohorts, with split protocol.}
\label{tab:prior}
\begin{tabular}{lllll}
\thickhline
study (best model) & cohort & RMSE@30 (mg/dL) & split & sd over \\
\hline
GlucoFM-Bench \cite{lu2026glucofmbench} (LSTM) & HUPA-UCM & 16.71 (4.82) & by-time & patients \\
GlucoFM-Bench \cite{lu2026glucofmbench} (LSTM) & T1D-UOM  & 20.81 (4.26) & by-time & patients \\
Zhu et al.\ \cite{zhu2023fcnn} (FCNN)          & ARISES   & 20.23 (3.38) & by-time & patients \\
Zhu et al.\ \cite{zhu2023fcnn} (FCNN)          & ABC4D    & 20.25 (2.60) & by-time & patients \\
\hline
Ours (MLDG, global)            & HUPA-UCM & 18.46 (0.15) & \emph{by-patient} & seeds \\
Ours (MLDG, global)            & ABC4D    & 19.68 & \emph{by-patient} & seeds \\
Ours (MLDG, global)            & ARISES   & 22.30 & \emph{by-patient} & seeds \\
Ours (MLDG, global)            & T1D-UOM  & 19.53 & \emph{by-patient} & seeds \\
\thickhline
\end{tabular}
\begin{center}
{\small A \emph{by-time} split tests on later windows of patients seen during
training; a \emph{by-patient} split tests on entirely unseen patients and is the
harder, deployment-relevant task. External sd is across patients; ours is across
5 seeds. Numbers are indicative, not head-to-head, because the split (and, for
Zhu et al., the input features) differ. Other papers on these cohorts report
non-comparable metrics or horizons: Tan \& McBeth \cite{tan2026uncertainty}
(MARD only), AEGIS \cite{pansheriya2026aegis} (15-min horizon, mmol/L),
Manchanda et al.\ \cite{manchanda2025lstm} (window-averaged RMSE), and Kolev
et al.\ \cite{kolev2026benchmark} (synthetic data only).}
\end{center}
\end{table}

\paragraph{Limitations.} Several constraints bound these conclusions. \emph{(i)
Evaluation metric.} We report a single aggregate error (RMSE at 30 min). This is
behind the norm of the comparator literature, which reports clinical error grids
(Clarke/Parkes), MARD, or hypo-/hyperglycaemia detection; aggregate RMSE is
dominated by the euglycaemic range and can be near-blind to the safety-critical
hypoglycaemic tail. Our model also emits a point forecast with no calibrated
predictive uncertainty. Whether federation's small advantage survives on
clinically stratified and event-detection metrics is untested and is the most
important gap for a clinical reading of this work. \emph{(ii) A single OOD cohort.}
Our transfer claim rests on one held-out cohort (ReplaceBG), which is itself a
large, fairly conventional adult US T1D cohort; a leave-one-cohort-out design
over the five cohorts would give several OOD tests and is the highest-value next
experiment. \emph{(iii) Statistical power.} Margins among federated strategies
($\approx 0.1$ mg/dL) are within seed standard deviation, some cells use 4 rather
than 5 seeds, and we ran no significance test, so we do not rank the federated
strategies. Seed sd also reflects one fixed patient split and understates
patient-sampling uncertainty. \emph{(iv) Baselines.} We include no persistence or
simple learned baseline in the main table, so we cannot fully attribute accuracy
to the transformer versus the training scheme; FedProx used a single untuned
$\mu$, and APFL may be under-tuned. \emph{(v) Horizon.} We report only 30 min,
though the model forecasts to 60 min. \emph{(vi) Privacy.} Raw data never moves,
but the prototype provides no formal guarantee: it uses plain HTTP without
encryption, secure aggregation, or differential privacy, and the exchanged
updates could in principle be attacked. \emph{(vii) Population.} All five cohorts
are adult T1D, leaving paediatric and type-2 transfer untested, and we scope the
benchmark to meta-learning and personalisation, leaving representation-alignment
methods to future work.

\paragraph{Future work.} The next steps follow directly from these limitations.
Most immediately: report clinical error-grid, MARD, and hypo-/hyperglycaemia
detection metrics (with predictive uncertainty) to test whether federation's
advantage holds where safety matters; add persistence and simple learned
baselines and paired significance tests; and run a leave-one-cohort-out
evaluation for several OOD estimates rather than one. Beyond that: extend to the
60-minute horizon and to event-aware inputs such as insulin and carbohydrate
information; add secure aggregation and differential privacy so the
no-raw-data-movement property is backed by a formal privacy guarantee; and scale
the prototype toward a genuinely multi-site, multi-country deployment with
reported communication, latency, and failure behaviour. Finally, combining the
generalisation of a federated model with per-site finetuning --- ideally trained
jointly, and studied as a function of how much local data a new site has --- and
continual federation as populations and devices drift over time are natural
extensions this framework is positioned to support.


% ===========================================================================
% Compile references.bib with the plos2025.bst style (ships with PLOS template).
\bibliography{references}

\end{document}
